\documentclass{beamer}
\beamertemplatenavigationsymbolsempty
\usepackage{tikz}
\begin{document}



\begin{frame}{Review: Chromatic polynomial of $C_4$ and a Lemma}  
\begin{lemma}Let $x$ and $y$ be two non-adjacent vertices in $G$.  Then 
$$P_G(k)=P_{G_{+xy}}(k)+P_{G_{x=y}}(k)$$
\end{lemma}

  \begin{columns}
     \begin{column}{.5\textwidth}
\begin{block}{1 and 3 have different colours}

  \centering
  \begin{tikzpicture}[scale=2]
\begin{scope}[every node/.style={circle, draw}]
\node (1) at (0,1) {$1$};
\node (2) at (1,1) {$2$};
\node (3) at (1,0) {$3$};
\node (4) at (0,0) {$4$};
\draw (1)--(2)--(3)--(4)--(1)--(3);
\end{scope}
\end{tikzpicture}
$k(k-1)(k-2)^2$
\end{block}
     \end{column}
 \begin{column}{.5\textwidth}
\begin{block}{1 and 3 have the same colour}

  \centering
  \begin{tikzpicture}[scale=2]
\begin{scope}[every node/.style={circle, draw}]
\node (1) at (.5,.5) {$*$};
\node (2) at (1,1) {$2$};
\node (4) at (0,0) {$4$};
\draw (2) to[bend right] (1);
\draw (1) to[bend right] (4);
\draw (4) to[bend right] (1);
\draw (1) to[bend right] (2);
\end{scope}
  \end{tikzpicture}

$k(k-1)^2$
\end{block}
     \end{column}
  \end{columns}
$$P_{C_4}(k)=k(k-1)(k-2)^2+k(k-1)^2=k(k-1)(k^2-3k+3)$$  
  \end{frame}
\begin{frame}{Deletion-Contraction: restructuring the lemma}
The lemmais useful for calculating, but awkward for induction:
\begin{itemize}
    \item $G_{+xy}$ has an extra edge; "bigger"
    \item $G_{x=y}$ has fewer vertices: "smaller"
\end{itemize}
Rewrite so $H\cong G_{+xy}$ is the "starting" graph.
\begin{lemma}[Deletion-Contraction]
Let $H$ be a graph, and let $e$ be an edge in $H$.  Then
$$P_H(k)=P_{H\setminus e}(k)-P_{H/e}(k)$$
\end{lemma}  
The edge $e$ is between $xy$
$$H=G+xy,\qquad H\setminus e=G,\qquad H/e=G_{x=y}$$

\end{frame}  



\begin{frame}{Chromatic polynomial is a polynomial}
\begin{theorem}
Let $G$ be a simple graph. Then $P_G(k)$ is a polynomial in $k$.  

Moreover, if $G$ has $n$ vertices and $m$ edges, then 
$$P_G(k)=k^n-mk^{n-1}+\text{lower order terms}$$
\end{theorem}

\begin{block}{Proof idea:}
Induct on the number of edges using deletion-contraction.
\end{block}
\begin{block}{Base case(s): $m=0$, any $n$}
If $G$ has no edges and $n$ vertices, then $G=E_n$ empty graph.
$P_{E_n}=k^n$ is a polynomial of the right form.
\end{block}

\end{frame}
\begin{frame}{Inductive step}
Assume that $G$ has $m>0$ edges and $n$ vertices, and that for any graph $H$ with $\ell<m$ edges and $p$ vertices, we have $P_H(k)=k^p-\ell k^{p-1}+\cdots$.

\begin{block}{Let $e\in G$ be any edge:}
\begin{itemize}
    \item $G\setminus e$ has $n$ vertices and $m-1$ edges
    \item $G/e$ has $n-1$ vertices and \emph{at most} $m-1$ edges
\end{itemize}
So by the inductive hypothesis, theorem holds for $G\setminus e$ and $G/e$
\end{block}
\begin{block}{So applying Deletion-Contraction:}\begin{align*}
    P_G(k) &= P_{G\setminus e}(k)-P_{G/e}(k) \\
    &= \left(k^n-(m-1)k^{n-1} +\cdots\right) -\left(k^{n-1}-\cdots\right) \\
    &= k^n-mk^{n-1}+\cdots
\end{align*}
Which is what we needed to show. \quad$\square$
\end{block}

\end{frame}
\begin{frame}{Odds and Ends}
  \begin{block}{Deletion-Contraction as an algorithm}
    \begin{itemize}
    \item Can always find $P_G(x)$ by iterating deletion-contraction
    \item In practise, often faster to add edges
      \end{itemize}
  \end{block}

  \begin{block}{Information in $P_G(k)$}
    \begin{itemize}
    \item Number of vertices is the degree
    \item Number of edges is negative the coefficient of next highest term
      \item $\chi(G)$ is the lowest $k$ with $P_G(k)\neq 0$.
\end{itemize}

    \end{block}





\end{frame}


\end{document}
 Binary file addedBIN +119 KB 
Slides/Lecture20.pdf
Binary file not shown.
 173 changes: 173 additions & 0 deletions173  
Slides/Lecture20.tex
@@ -0,0 +1,173 @@
\documentclass{beamer}
\beamertemplatenavigationsymbolsempty
\usepackage{tikz}
\begin{document}

\begin{frame}{More techniques for computing chromatic polynomials}
  \begin{block}{Main goal: two new bits}
    \begin{itemize}
    \item Use induction to calculate $P_G(k)$ for a family of graphs.
      We'll do $C_n$
    \item Gluing formulas for graphs that are nearly disconnected
 \end{itemize}
    \begin{tikzpicture}[every node/.style={circle, fill, draw}]
      \begin{scope}[yshift=.5cm]
      \node (1) at (0:1) {};
      \node (2) at (72:1) {};
      \node (3) at (144:1) {};
      \node (4) at (216:1) {};
      \node (5) at (288:1) {};
      \node (6) at (2,1) {};
      \node (7) at (3, 0) {};
      \node (8) at (2,-1) {};
      \draw (1)--(2)--(3)--(4)--(5)--(1)--(6)--(7)--(8)--(1);

        \end{scope}


      \begin{scope}[xshift=5cm]
      \node (1) at (0,0) {};
      \node (2) at (0,1) {};
      \node (3) at (1,1) {};
      \node (4) at (1,0) {};
      \node (5) at (2,1.5) {};
      \node (6) at (3,.5) {};
      \node (7) at (2, -.5) {};
      \draw (3)--(2)--(1)--(4)--(3)--(5)--(6)--(7)--(4);
\end{scope}
      \end{tikzpicture}

    The graphs above glued from $C_4$ and $C_5$.
\end{block}
\begin{block}{If we have time:}
  Compute $P_G(k)$ for a relatively complicated graph $G$.
\end{block}
\end{frame}    



\begin{frame}{Calculating the chromatic polynomial of $C_n$}
  Let $e$ be any edge of $C_n$.  Then:
  \begin{itemize}
  \item $C_n/e \cong C_{n-1}$
  \item $C_n\setminus e=P_n$, a tree, so $P_{P_n}(k)=k(k-1)^{n-1}$
  \end{itemize}

  So we should be able to find $P_{C_n}(k)$ using induction, but need to ``guess'' the formula first.
 \begin{align*}
    P_4(k)= & k^4-4k^3+6k^2-3k \\
P_5(k)= & k^5-5k^4+10k^3-10k^2+4k \\
P_6(k)= & k^6-6k^5+15k^4-20k^3+15k^3-5k \\
P_7(k)= & k^7-7k^6+21k^5-35k^4+35k^3-21k^2+6k
    \end{align*}
\begin{block}{Looks like:}
$$P_n(k)=(k-1)^n+(-1)^n(k-1)$$
  \end{block}
\end{frame}

\begin{frame}{Inductive proof that $P_{C_n}(k)=(k-1)^n+(-1)^n(k-1)$}
  \begin{block}{Base case: $n=3$}
Plug in $n=3$, get $k(k-1)(k-2)=P_{C_3}(k)$.
  \end{block}

  \begin{block}{Inductive step:}
   Get to assume: $P_{C_{n-1}}(k)=(k-1)^{n-1}+(-1)^{n-1}(k-1)$

    \begin{itemize}
      \item $C_n\setminus e=P_n$, a tree, so $P_{C_{n-1}\setminus e}=k(k-1)^{n-1}$
      \item $C_n/e=C_{n-1}$, so $P_{C_n/e}(k)=(k-1)^{n-1}+(-1)^n(k-1)$.
        \end{itemize}
\end{block}
    \begin{block}{Plugging into deletion-contraction:}   \begin{align*}
      P_{C_n}(k)&=P_{C_n\setminus e}(k)-P_{C_n/e}(k) \\
      &=k(k-1)^{n-1}-\left[(k-1)^{n-1}+(-1)^{n-1}(k-1)\right] \\
      &=k(k-1)^{n-1}-(k-1)^{n-1}-(-1)^{n-1}(k-1) \\
      &=(k-1)^{n-1}\left[k-1\right] +(-1)^n(k-1) \\
      &=(k-1)^n+(-1)^n(k-1)
    \end{align*}
    \end{block}

\end{frame}

\begin{frame}{Gluing formulas: intro}
  \begin{lemma} If $G$ is the disjoint union of $G_1$ and $G_2$, then
    $$P_G(k)=P_{G_1}(k)P_{G_2}(k)$$
    \end{lemma}
\begin{proof} Colouring $G$ is exactly the same as colouring $G_1$ and $G_2$ independently.
\end{proof}


\begin{block}{Gluing formulas: when $G$ isn't \emph{quite} a disjoint union}
  Idea: Colour $G_1$, then extend to a colouring of $G_2$.
\end{block}
\end{frame}
\begin{frame}{Gluing formulas: statements}
    \begin{lemma}If $G$ is made by gluing $G_1$ and $G_2$ along a vertex $v$, then:

      $$P_G(k)=\frac{1}{k}P_{G_1}(k)P_{G_2}(k)$$
      \end{lemma}




    \begin{proof}First colour $G_1$ in any of the $P_{G_1}(k)$ ways.  Now, vertex $v$ of $G_2$ is already coloured, but none of the rest.  Since the colours are interchangable, exactly $1/k$ of the ways of colouring $G_2$ will have the right colour at $v$.
\end{proof}
    \begin{lemma}
      If $G$ is made by gluing $G_1$ and $G_2$ along an edge $e$, then $$P_G(k)=\frac{1}{k(k-1)} P_{G_1}(k)P_{G_2}(k)$$
\end{lemma}



\end{frame}


\begin{frame}{Find $P_G(k)$ for the following graphs}
  \centering
  \begin{tikzpicture}[every node/.style={circle, fill, draw}]
    \begin{scope}
      \node (1) at (1,1) {};
      \node (2) at (0,0) {};
      \node (3) at (1,0) {};
      \node (4) at (2,0) {};
      \node (5) at (1,-1) {};
      \draw (1)--(2)--(5)--(3)--(1)--(4)--(5);
    \end{scope}

      \begin{scope}[xshift=4cm]
      \node (1) at (1,1) {};
      \node (2) at (0,0) {};
      \node (3) at (1,0) {};
      \node (4) at (2,0) {};
      \node (5) at (1,-1) {};
      \draw (1)--(2)--(5)--(3)--(1)--(4)--(5);
\draw (2)--(3);
      \end{scope}

        \begin{scope}[xshift=8cm]
      \node (1) at (1,1) {};
      \node (2) at (0,0) {};
      \node (3) at (1,0) {};
      \node (4) at (2,0) {};
      \node (5) at (1,-1) {};
      \draw (1)--(2)--(5)--(3)--(1)--(4)--(5);
      \draw (2)--(3)--(4);
        \end{scope}

  \begin{scope}[xshift=4cm, yshift=-4cm]
      \node (1) at (1,1) {};
      \node (2) at (0,0) {};
      \node (3) at (1,0) {};
      \node (4) at (2,0) {};
      \node (5) at (1,-1) {};
      \draw (1)--(2)--(5)--(3)--(1)--(4)--(5);
      \draw (2)--(3)--(4);
      \draw (2) to[out=90, in=180] (1,1.5) to[out=0,in=90] (4);

  \end{scope}

    \end{tikzpicture}


  \end{frame}

\end{document}
