\documentclass{beamer}
\beamertemplatenavigationsymbolsempty
\usepackage{amsmath, amssymb, hyperref, graphics}
\usepackage{tikz}
\usetikzlibrary{arrows}


\title{Graph Theory Lecture 9}

\begin{document}
\begin{frame}{Last time: cheapest spanning trees}
  \begin{block}{Two ways: both greedy algorithms}

    \begin{itemize}
    \item Kruskal: add the cheapest edge that doesn't make a loop
    \item Prim: start at $v$, add cheapest edge to a new vertex
    \end{itemize}
  \end{block}

  \begin{block}{The tricky part:}
    The algorithms are easy to run, but it's not immediately clear that they actually find the cheapest spanning tree.  The proof that they do will not appear on the exam.
    \end{block}

  \begin{block}{Today: shortest and longest paths; Travelling Salesperson}
    \end{block}

\end{frame}

\begin{frame}{Directed graphs}
  \begin{definition} A \emph{directed graph} is one where each edge has a chosen starting and ending point, usually indicated with arrows.
 \end{definition}
\begin{center}
\begin{tikzpicture}[scale=.8,->,>=stealth', shorten >=1pt,auto,node distance=4cm,
                thick,main node/.style={circle,draw,font=\Large\bfseries}]

  \node[main node] (1) {A};
  \node[main node] (2) [below left of=1] {B};
  \node[main node] (3) [below right of=1] {C};

  \path
    (1) edge [bend right] node[above] {4} (2)
    (2) edge node [below]{7} (1)
    (3) edge [loop below] node {8} (3)
        edge node[right] {2} (1)
        edge node[below] {3} (2);      
\end{tikzpicture}
\end{center}


  
  \begin{block}{Walks in directed graphs:}
You can only travel the edge in the direction the arrow shows.
    \end{block}
\end{frame}

\begin{frame}{Dijkstra's algorithm for shortest path}
  \begin{block}{Input:} A weighted (possibly directed) graph $G$ and starting vertex $v\in G$
  \end{block}
  \begin{block}{Output:} For every vertex $w\in G$ a list of all shortest paths from $v$ to $w$
  \end{block}
  
 \begin{block}{Initialize:}
   From starting vertex $v$ list every edge out of $v$ as a poential shortest path to corresponding vertex $w$
   \end{block}
       \begin{block}{Iterate:}
 \begin{itemize}
 \item Make whichever shortest path $w$ 


 \item Update list of potential paths by adding edges out of $w$ to the shortest paths to $w$ and checking if they're cheaper than known paths
   \end{itemize}
 \end{block}
  \end{frame}

\begin{frame}{Example graph from the 2008 Exam}
Find all shortest paths from $S$ to $T$.
\begin{center}
\begin{tikzpicture}[scale=1.3, ->,>=stealth', shorten >=1pt,auto,node distance=4cm,
                thick,main node/.style={circle,draw}]

  \node[main node] (S) at (0,1) {S};
  \node[main node] (A) at (1,2) {A};
  \node[main node] (B) at (1,0) {B};
  \node[main node] (C) at (3,2) {C};
  \node[main node] (D) at (3,0) {D};
  \node[main node] (E) at (5,2) {E};
  \node[main node] (F) at (5,0) {F};
  \node[main node] (T) at (6,1) {T};
  
  \path
    (S) edge node {11} (A)
      edge node[left] {7} (B)
      (B) edge node[below] {10} (D)
      edge node[right] {3} (A)
      (A)       edge node[above] {10} (C)
      edge node[right] {6} (D)
      (C)       edge node[above] {10} (E)
      edge node[right] {5} (F)
      (D)    edge node[right] {4} (C)
      edge node[below] {9} (F)
      (E)      edge node[right] {7} (T)
      (F)      edge node[right] {11} (T)
            edge node[right] {3} (E);
\end{tikzpicture}
\end{center}

\begin{block}{Add on:}
\begin{itemize}
\item  Which edges if made a little longer would make the distance from $S$ to $T$ longer?
\item  Which edges if made a little shorter would make the distance from $S$ to $T$ shorter?
\end{itemize}
\end{block}

  \end{frame}


\begin{frame}{The main idea of why Dijkstra's works:}


Suppose we're at the step were Dijkstra's decides the cheapest path to $w$. 
\begin{block}{Then:}
A cheaper path to $w$ would have to go through a vertex $u$ we haven't found the cheapest path to yet.
  \end{block}

  \begin{block}{But!}
Even \emph{getting} to $u$ costs more than our cheapest path to $w$.
    \end{block}

  \begin{block}{An observation:}
Dijkstra's algorithm depends on the edge weights being non-negative!
  \end{block}
  
  \end{frame}

\begin{frame}{Culture: performance of Dijkstra's algorithm}
  \begin{block}{In a limited sense, Dijkstra's algorithm is optimal:}
If \emph{all we know} is that we have a weighted graph, then you can't do better than Dijkstra's algorithm.
    \end{block}
  \begin{block}{In practise, often not very good:}
 When finding path from Sheffield to Edinburgh, Dijkstra's algorithm explores every street in London.
  \end{block}
  \begin{block}{Real world maps have extra information:}
    It's easy to calculate the distance between two points as the crow flies, and we know the driving distance has to be at least that large.  
\end{block}  
  \begin{block}{The A* algorithm avoids searching London:}
    Supposes we have an easy to calculate ``heuristic distance'' $h(v,w)$, that is a lower bound for the actual distance $d(v,w)$. 
    \end{block}
\end{frame}


\begin{frame}{Finding the longest path}
  \begin{block}{Scheduling a large problem with many parts}
For example, building a house.
    \begin{itemize}
    \item Some can be done at same time: (finishing interior rooms)
    \item Some need to be done in order: (foundation before walls)
    \item How early could whole project be finished?
    \end{itemize}
  \end{block}
  \begin{block}{Solution: longest path}
    \begin{itemize}
    \item  Encode tasks as edges in directed weighted graph        
    \item Edge $e$ follows edge $f$ if task $f$ requires task $e$ 
    \item Length of longest path is the shortest time to complete project 
    \end{itemize}
\end{block}
Building the directed graph from a list of tasks and dependencies can require a few tricks and won't be tested.

\end{frame}

\begin{frame}{Longest path might not exist:}

  \begin{block}{A necessary assumption:}
    If the graph has a directed cycle, we could get an infinitely long path by repeating graph over and over again.  
  \end{block}

  \begin{definition}A directed graph $G$ is \emph{acyclic} if it has no directed cycles.
  \end{definition}

  \begin{block}{Graphs in scheduling applications are acyclic:}
 Otherwise we'd have a cycle of tasks that all depend on each other and we could never start the project!  
\end{block}

  \begin{block}{Ordering vertices / ``topological sort''}
    If $G$ is acyclic it's easy to order the vertices of $G$ so that if there's an edge from $v$ to $w$, then $w$ comes after $v$.
\end{block}
\end{frame}

\begin{frame}{The longest path algorithm:}
\begin{block}{Initializing:}
  \begin{itemize}
    \item Topologically sort the vertices of $G$ 
    \item List every edge of starting vertex as potential longest path
  \end{itemize}
\end{block}

\begin{block}{Iterate:}
  \begin{itemize}
  \item Make the potential longest path to the first vertex $w$  on the list permament
  \item Update the list of potential longest paths adding edges out of $w$ to longest paths to $w$ and seeing if they create new longest paths
  \end{itemize}
  \end{block}
  
  

  
  \end{frame}


\begin{frame}{Example graph from the 2008 Exam}
Find all longest paths from $S$ to $T$.
\begin{center}
\begin{tikzpicture}[scale=1.3, ->,>=stealth', shorten >=1pt,auto,node distance=4cm,
                thick,main node/.style={circle,draw}]

  \node[main node] (S) at (0,1) {S};
  \node[main node] (A) at (1,2) {A};
  \node[main node] (B) at (1,0) {B};
  \node[main node] (C) at (3,2) {C};
  \node[main node] (D) at (3,0) {D};
  \node[main node] (E) at (5,2) {E};
  \node[main node] (F) at (5,0) {F};
  \node[main node] (T) at (6,1) {T};
  
  \path
    (S) edge node {11} (A)
      edge node[left] {7} (B)
      (B) edge node[below] {10} (D)
      edge node[right] {3} (A)
      (A)       edge node[above] {10} (C)
      edge node[right] {6} (D)
      (C)       edge node[above] {10} (E)
      edge node[right] {5} (F)
      (D)    edge node[right] {4} (C)
      edge node[below] {9} (F)
      (E)      edge node[right] {7} (T)
      (F)      edge node[right] {11} (T)
            edge node[right] {3} (E);
\end{tikzpicture}
\end{center}

\begin{block}{Add on:}
\begin{itemize}
\item  Which edges if made a little longer would make the longest path from $S$ to $T$ longer?
\item  Which edges if made a little shorter would make the longest path from $S$ to $T$ shorter?
\end{itemize}
\end{block}

  \end{frame}
  


\end{document}
